title: "Clustering_Env"
author: "Connor Hughes"
date: "2023-03-29"
output: html_document
---


Load required packages
```{r Load Packages, warning=FALSE}
# Check if packages are installed, if not, install them
packages <- c("dplyr", "raster", "ggplot2", "mapdata", "cluster", "factoextra", "dendextend", "reshape2", "cowplot")
for (package in packages) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
}

# Load the packages
library(dplyr)      # for data manipulation
library(raster)     # for handling spatial data
library(ggplot2)    # for data visualization
library(mapdata)    # for generating maps
library(cluster)    # for clustering analysis
library(factoextra) # for clustering visualizations
library(dendextend) # for modifying dendrograms
library(reshape2)   # for data transformation
library(cowplot)    # for plotting multiple plots together

```

Data Preparation
Load data and subset only complete cases
```{r Data Preparation}
#Clear environment 
rm(list=ls())

#Choose a state and type it below!

State <- zz

My_State <- read.csv(paste0("./SamplePts/",State,"_Points.csv")) # load data from CSV file, choose your state here
My_State <- subset(My_State, complete.cases(My_State) == T) # subset only complete cases (rows without any missing values)
My_State$Ecoregion <- My_State$na_l2name # create a new column "Ecoregion" and assign values from "na_l2name" column
```


Visualize the Distribution of your Sampled Points

```{r Visualize the Distribution}
#Load in a map of the US
us_map <- map_data("state")

#Plot the sampled points on this map of the US
ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = My_State, aes(x = Longitude, y = Latitude), 
             color = "firebrick", size = .5, stroke = 1.5) +
  labs(x = "Longitude", y = "Latitude") +
  coord_quickmap(xlim = c(min(My_State$Longitude)-.5, max(My_State$Longitude)+.5), 
                  ylim = c(min(My_State$Latitude)-.5, max(My_State$Latitude))+.5) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")
```

Visualize the Distribution of your points in Envrionmental Space

```{r E Space}
# Plot Bio 1 (Annual Temprature) and Bio 12 (Annual Precipitation) of these points
ggplot(My_State, aes(My_State$bio01/10, My_State$bio12/10)) +
  geom_point() +
  labs(y = "Precipitation", x = "Temprature (C)")
```
Isolate and Scale your Environmental Variables

```{r Isolate and Scale}
#Read in collumn names of 'My State'
colnames(My_State)
#Select only those we're interested in clustering by, in this case Bio01 and Bio12
My_State_bios <- dplyr::select(My_State, bio01, bio12)
#Scale these variables so that they have comparable values
My_State_scaled <- scale(My_State_bios)
```
Test which linkage method is best for your data set 

```{r Linkage Method}
#define linkage methods
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

#function to compute agglomerative coefficient (How strong are your clusters)
ac <- function(x) {
  agnes(My_State_scaled, method = x)$ac
}

#calculate agglomerative coefficient for each clustering linkage method
sapply(m, ac)
```

Choose an number of clusters for your data set - use these tests to help you do so

```{r Cluster Number, message=F, warning=F}
### Elbow method
fviz_nbclust(My_State_scaled, FUN = hcut, method = "wss", print.summary = T)

### Silhouette Method
fviz_nbclust(My_State_scaled, FUN = hcut, method = "silhouette", print.summary = T)
```
You can also visually assess your clusters to choose the number of groups you'd like to naturally create
```{r Test Dendrogram}
#compute distance matrix
d <- dist(My_State_scaled, method = "euclidean")

#perform hierarchical clustering using the method you previously selected
method <- "ward.D2"
prelim_clust <- hclust(d, method = method)

avg_dend_obj <- as.dendrogram(prelim_clust, labs = F)  %>% set("labels", "")
plot(avg_dend_obj)
```
Now, create your clusters
```{r Full Dendrogram}
#compute distance matrix
d <- dist(My_State_scaled, method = "euclidean")

#perform hierarchical clustering using the method you previously selected
method <- "ward.D2"
final_clust <- hclust(d, method = method)

#cut the dendrogram into the apropriate number of clusters you previously decided (Based off the previous two code blocks)
k = #
groups <- cutree(final_clust, k=k)

#Plot your dendrogram!
avg_col_dend <- color_branches(final_clust, k = k)  %>% set("labels", "")
plot(avg_col_dend)
```
Visualize some of the main differences between your groups
```{r Visualize Cluster}
#Append your new groupings to the primary data frame
My_State_scaled <- as.data.frame(My_State_scaled)
My_State_scaled$Cluster <- groups
My_State$Cluster <- groups

#Plot your clusters in environmental space
ggplot(My_State_scaled, aes(bio01, bio12, color = as.character(Cluster))) +
  geom_point() +
  labs(x = "Precipitation", y = "Temprature", color = "Cluster")

#Flatten your dataframe to allow the creation of the next plot
Cluster_Info <- melt(My_State_scaled, by=list(cluster=My_State_scaled$Cluster), id.vars = "Cluster")

#Make a bar plot visualizing the differences in your clusters
ggplot(Cluster_Info, aes(x = variable, y = as.numeric(value), fill = as.character(Cluster))) +
  geom_boxplot() +
  labs(x = "Bioclim Variable", y = "Value (SD from Mean)", fill = "Cluster") +
  coord_flip()
```
Plot your clusters in Geographic Space
```{r Plot Clusters in G Space}
ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = My_State, aes(x = Longitude, y = Latitude, fill = as.character(Cluster)), size = 3, shape = 21, stroke = 1.5) +
  labs( 
       x = "Longitude", y = "Latitude", fill = "Cluster") +
  coord_quickmap(xlim = c(min(My_State$Longitude)-.5, max(My_State$Longitude)+.5), 
                  ylim = c(min(My_State$Latitude)-.5, max(My_State$Latitude))+.5) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom")
```
Nice work! You just designated regions of this state based on environmental variables.

~~~

States already have some system of labeling different environmental provinces- these are more complex and often take into account more than just abiotic factors, but are a similar idea. How does our clustering match up against the states ecoregions? 


Lets see!
```{r Plot Ecoregions}
#Append ecoregion to your dataframe
My_State_scaled$Ecoregion <- My_State$Ecoregion
My_State$Cluster <- groups

#Plot the ecoregions in the same environmental space we used to create our clusters
ggplot(My_State_scaled, aes(bio01, bio12, color = as.character(Ecoregion))) +
  geom_point() +
  labs(x = "Precipitation", y = "Temprature", color = "Ecoregion")

#Make a map of the ecoregions by point
ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = My_State, aes(x = Longitude, y = Latitude, fill = as.factor(Ecoregion)), size = 3, shape = 21, stroke = 1.5) +
  labs( 
       x = "Longitude", y = "Latitude", fill = "Ecoregion") +
  coord_quickmap(xlim = c(min(My_State$Longitude)-.5, max(My_State$Longitude)+.5), 
                  ylim = c(min(My_State$Latitude)-.5, max(My_State$Latitude))+.5) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom",
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 4))

```
How does this compare to your grouping? Are states ecoregions primarily grouped by temprature and precipitation? or are there likely other factors at play?

```{r Clusters vs. Ecoregions}
# create a data frame of Clusters vs. Ecoregions
df <- data.frame(Cluster = My_State_scaled$Cluster,
                 Ecoregion = My_State_scaled$Ecoregion)

# Plot this as a bar graph
ggplot(df, aes(x = as.factor(Cluster), fill = Ecoregion)) +
  geom_bar(position = "dodge") +
  labs(x = "Cluster", y = "Count", fill = "Ecoregion") +
  theme_minimal() +
  geom_vline(xintercept = c(seq(1.5, k, 1)), linetype = "dashed")

```

Lets look at ecoregions and our groupings side by side
```{r}
p1 <-ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = My_State, aes(x = Longitude, y = Latitude, fill = as.character(Cluster)), size = 3, shape = 21, stroke = 1.5) +
  labs( 
       x = "Longitude", y = "Latitude", fill = "Cluster") +
  coord_quickmap(xlim = c(min(My_State$Longitude)-.5, max(My_State$Longitude)+.5), 
                  ylim = c(min(My_State$Latitude)-.5, max(My_State$Latitude))+.5) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")

p2 <- ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = My_State, aes(x = Longitude, y = Latitude, fill = as.factor(Ecoregion)), size = 3, shape = 21, stroke = 1.5) +
  labs( 
       x = "Longitude", y = "Latitude", fill = "Ecoregion") +
  coord_quickmap(xlim = c(min(My_State$Longitude)-.5, max(My_State$Longitude)+.5), 
                  ylim = c(min(My_State$Latitude)-.5, max(My_State$Latitude))+.5) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none",
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 4))

plot_grid(p1, p2)
```